---
title: "Getting Started with maxentcpp"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This article walks through a complete species distribution modelling workflow
using `maxentcpp`.
We model the distribution of the Emerald-bellied Hummingbird
(*Abeillia abeillei*) using two bioclimatic variables bundled with the package.

## Installation

```r
# From CRAN (when available)
install.packages("maxentcpp")

# Or from GitHub
remotes::install_github("alrobles/maxentcpp")
```

## Load the Package and Data

`maxentcpp` ships with a cropped bioclimatic raster stack (bio1: Annual Mean
Temperature, bio12: Annual Precipitation) and 73 GBIF occurrence records for
*Abeillia abeillei*.

```{r load-data}
library(maxentcpp)
library(terra)

# Load environmental layers
stack_path      <- system.file("extdata", "stack_1_12_crop.rds",
                               package = "maxentcpp")
example_rasters <- terra::unwrap(readRDS(stack_path))
bio1  <- example_rasters[[1]]
bio12 <- example_rasters[[2]]

# Convert to maxentcpp grid objects
g_bio1  <- maxent_grid_from_terra(bio1)
g_bio12 <- maxent_grid_from_terra(bio12)

# Load occurrence records
data(example_occ_df)
head(example_occ_df)
```

## Prepare Spatial References

Create a grid dimension object that matches the environmental layers,
then convert occurrence lon/lat coordinates to grid row/column indices:

```{r prepare-spatial}
info <- maxent_grid_info(g_bio1)
dim  <- maxent_dimension(nrows    = info$nrows,
                         ncols    = info$ncols,
                         xll      = info$xll,
                         yll      = info$yll,
                         cellsize = info$cellsize)

occ <- maxent_read_occurrences(example_occ_df, dim,
                               lon_col = "long",
                               lat_col = "lat")

cat("Presence points:", length(occ$indices), "\n")
```

## Sample Background Points

Maxent contrasts presence locations against a random sample of background
points.
The default of 10 000 background points is standard practice:

```{r background}
bg <- maxent_background_indices(g_bio1, n = 10000, seed = 42)
cat("Background points:", length(bg$indices), "\n")
```

## Extract Environmental Values and Generate Features

Extract the environmental variable values at all point locations
(background + presence) and generate the feature transformations
(linear, quadratic, and hinge):

```{r features}
all_rows <- c(bg$rows, occ$rows)
all_cols <- c(bg$cols, occ$cols)
n_total  <- length(all_rows)

# 0-based indices of the presence samples within the combined vector
sample_indices <- seq(length(bg$rows), n_total - 1L)

# Extract environmental values at all locations
bio1_vals  <- sapply(seq_along(all_rows), function(i)
    grid_get_value(g_bio1,  all_rows[i], all_cols[i]))
bio12_vals <- sapply(seq_along(all_rows), function(i)
    grid_get_value(g_bio12, all_rows[i], all_cols[i]))

env_data <- list(bio1 = bio1_vals, bio12 = bio12_vals)

features <- maxent_generate_features(env_data,
                                     types    = c("linear", "quadratic", "hinge"),
                                     n_hinges = 15)
cat("Generated", length(features), "features\n")
```

## Train the Model

Create a `FeaturedSpace` object and train the Maxent model using
sequential coordinate ascent --- the same algorithm used by Java Maxent:

```{r train}
fs     <- maxent_featured_space(n_total, as.integer(sample_indices), features)
result <- maxent_fit(fs,
                     max_iter        = 500,
                     convergence     = 1e-5,
                     beta_multiplier = 1.0)

cat("Converged:", result$converged, "\n")
cat("Iterations:", result$iterations, "\n")
cat("Entropy:", round(result$entropy, 4), "\n")
```

## Evaluate Model Performance

Compute the Area Under the ROC Curve (AUC) and other metrics by comparing
predictions at presence sites versus background sites:

```{r evaluate}
pres_preds <- maxent_extract_predictions_raw(
    fs, list(g_bio1, g_bio12), c("bio1", "bio12"),
    occ$rows, occ$cols)

bg_preds <- maxent_extract_predictions_raw(
    fs, list(g_bio1, g_bio12), c("bio1", "bio12"),
    bg$rows, bg$cols)

eval_result <- maxent_evaluate(pres_preds, bg_preds)
cat("AUC:", round(eval_result$auc, 4), "\n")
cat("Max Kappa:", round(eval_result$max_kappa, 4), "\n")
```

## Project onto the Landscape

Generate a spatial prediction map using the complementary log-log (cloglog)
transformation, which produces values in [0, 1] interpretable as
probability of presence:

```{r project, fig.alt = "Predicted habitat suitability map for Abeillia abeillei using cloglog output."}
pred_grid   <- maxent_project_cloglog(fs,
                                      list(g_bio1, g_bio12),
                                      c("bio1", "bio12"))
pred_raster <- maxent_grid_to_terra(pred_grid)

terra::plot(pred_raster,
            main = "Abeillia abeillei - Predicted Habitat Suitability",
            col  = hcl.colors(50, "YlOrRd", rev = TRUE))
```

## Variable Importance

Assess which environmental variables contribute most to the model:

```{r importance}
contrib <- maxent_percent_contribution(fs, c("bio1", "bio12"))
cat("Percent contribution:\n")
print(contrib)

perm_imp <- maxent_permutation_importance(
    fs, list(g_bio1, g_bio12), c("bio1", "bio12"),
    occ$rows, occ$cols,
    bg$rows,  bg$cols)
cat("\nPermutation importance:\n")
print(perm_imp)
```

## Response Curves

Visualise how the predicted suitability changes along each environmental
gradient:

```{r response-curves, fig.alt = "Response curves showing predicted suitability as a function of bio1 and bio12."}
par(mfrow = c(1, 2))

rc_bio1 <- maxent_response_curve(fs,
                                 list(g_bio1, g_bio12),
                                 c("bio1", "bio12"),
                                 var_index = 0L)
plot(rc_bio1$value, rc_bio1$prediction, type = "l", lwd = 2,
     xlab = "bio1 (Annual Mean Temperature)",
     ylab = "Predicted Suitability",
     main = "Response: bio1")

rc_bio12 <- maxent_response_curve(fs,
                                  list(g_bio1, g_bio12),
                                  c("bio1", "bio12"),
                                  var_index = 1L)
plot(rc_bio12$value, rc_bio12$prediction, type = "l", lwd = 2,
     xlab = "bio12 (Annual Precipitation)",
     ylab = "Predicted Suitability",
     main = "Response: bio12")

par(mfrow = c(1, 1))
```

## MESS: Multivariate Environmental Similarity Surface

Identify areas where the model is extrapolating beyond the training
environment.
Negative MESS values indicate novel environmental combinations:

```{r mess, fig.alt = "MESS map showing areas of environmental novelty relative to training data."}
mess_grid   <- maxent_mess(list(g_bio1, g_bio12), c("bio1", "bio12"),
                           occ$rows, occ$cols)
mess_raster <- maxent_grid_to_terra(mess_grid)

terra::plot(mess_raster,
            main = "MESS (negative = novel environment)",
            col  = hcl.colors(50, "RdBu"))
```

## Save and Load Models

Models are persisted as lambda files, fully compatible with Java Maxent:

```{r save-load}
lambdas_file <- file.path(tempdir(), "abeillei_model.lambdas")
maxent_save_lambdas(fs, lambdas_file)
cat("Model saved to:", lambdas_file, "\n")

# Load it back
fs_loaded <- maxent_load_lambdas(lambdas_file)
cat("Model loaded successfully\n")
```

## One-Click Workflow with `maxent_run()`

For a quick analysis, `maxent_run()` wraps the entire pipeline --- from
raw rasters and occurrence records to model outputs --- in a single call:
  
```r
result <- maxent_run(
    species    = "Abeillia_abeillei",
    env_grids  = list(bio1 = g_bio1, bio12 = g_bio12),
    occ_df     = example_occ_df,
    output_dir = tempdir(),
    lon_col    = "long",
    lat_col    = "lat"
)
```

This produces all standard outputs: lambda file, prediction PNG, response
curves, variable importance, omission/commission CSV, and
`maxentResults.csv`.

## Next Steps

- Read the [Why maxentcpp?](why_maxentcpp.html) article to understand the
  motivation behind the Java-to-C++ migration.
- See the [MaxEnt Features and Outputs](maxent_features_and_outputs.html)
  article for a deeper look at feature types and output transformations.
- Explore the [function reference](../reference/index.html) for the full
  API documentation.
