---
title: "MaxEnt Features and Outputs"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

This article explains the mathematical building blocks of the Maxent
algorithm as implemented in `maxentcpp`: what **features** are, how the
**optimizer** works, and what the different **output transformations** mean.

## The Maxent Principle

Maxent fits a probability distribution over geographic space by maximizing
entropy subject to constraints derived from the training data.
Given a set of environmental variables measured at presence locations, the
algorithm finds the distribution that:

1. Agrees with the observed data (sample expectations match model
   expectations for each feature).
2. Is as spread out (high entropy) as possible, avoiding unwarranted
   assumptions.

The result is a **Gibbs distribution**:

$$
P(x) = \frac{1}{Z} \exp\left(\sum_j \lambda_j f_j(x)\right)
$$

where $f_j$ are features (transformations of environmental variables),
$\lambda_j$ are learned weights, and $Z$ is the normalizing constant
(partition function).

## Feature Types

Features are transformations of the raw environmental variables.
Each feature type captures a different aspect of the species--environment
relationship.
`maxentcpp` implements all five feature types from Java Maxent 3.4.4:

### Linear Features

The simplest transformation: a linear rescaling to [0, 1].

$$
f_{\text{linear}}(v) = \frac{v - v_{\min}}{v_{\max} - v_{\min}}
$$

Linear features capture monotonic (directional) responses.
For example, a species that prefers warmer temperatures will have a
positive $\lambda$ for the linear feature of mean annual temperature.

```r
maxent_linear_feature(env_values, min_val, max_val)
```

### Quadratic Features

The square of the linear feature:

$$
f_{\text{quadratic}}(v) = f_{\text{linear}}(v)^2
$$

Quadratic features allow the model to capture **unimodal** (hump-shaped)
responses --- a species may prefer intermediate values of a variable,
declining at both extremes.

```r
maxent_quadratic_feature(env_values, min_val, max_val)
```

### Product Features

The product of two linear features from different variables:

$$
f_{\text{product}}(v_1, v_2) = f_{\text{linear}}(v_1) \times f_{\text{linear}}(v_2)
$$

Product features model **interactions** between environmental variables.
For example, a species might tolerate low precipitation only when
temperatures are also low.

```r
maxent_product_feature(env_values_1, env_values_2,
                       min1, max1, min2, max2)
```

### Threshold Features

A binary step function:

$$
f_{\text{threshold}}(v) =
\begin{cases}
1 & \text{if } v \geq t \\
0 & \text{otherwise}
\end{cases}
$$

Threshold features capture **hard limits** --- for example, a species that
cannot survive below a critical temperature.

```r
maxent_threshold_feature(env_values, threshold)
```

### Hinge Features

A piecewise linear function that is zero on one side and linear on the
other:

$$
f_{\text{hinge}}(v) =
\begin{cases}
0 & \text{if } v \leq \text{knot} \\
\frac{v - \text{knot}}{v_{\max} - \text{knot}} & \text{if } v > \text{knot}
\end{cases}
$$

(Reverse hinges ramp _down_ instead of up.)

Hinge features are the most expressive feature type in Maxent.
A set of evenly-spaced hinges can approximate any smooth response curve,
similar to a basis spline.
They are the default feature type for datasets with $\geq 15$ presence
records.

```r
maxent_hinge_feature(env_values, min_val, max_val, knot)
```

### Automatic Feature Generation

In practice, you rarely construct features manually.
`maxent_generate_features()` creates the full feature set from a list of
environmental variable vectors:

```{r generate-features}
library(maxentcpp)
library(terra)

stack_path      <- system.file("extdata", "stack_1_12_crop.rds",
                               package = "maxentcpp")
example_rasters <- terra::unwrap(readRDS(stack_path))
g_bio1  <- maxent_grid_from_terra(example_rasters[[1]])
g_bio12 <- maxent_grid_from_terra(example_rasters[[2]])

bg   <- maxent_background_indices(g_bio1, n = 10000, seed = 42)
data(example_occ_df)
info <- maxent_grid_info(g_bio1)
dim  <- maxent_dimension(info$nrows, info$ncols,
                         info$xll, info$yll, info$cellsize)
occ  <- maxent_read_occurrences(example_occ_df, dim,
                                lon_col = "long", lat_col = "lat")

all_rows <- c(bg$rows, occ$rows)
all_cols <- c(bg$cols, occ$cols)
bio1_vals  <- sapply(seq_along(all_rows), function(i)
    grid_get_value(g_bio1, all_rows[i], all_cols[i]))
bio12_vals <- sapply(seq_along(all_rows), function(i)
    grid_get_value(g_bio12, all_rows[i], all_cols[i]))

features <- maxent_generate_features(
    list(bio1 = bio1_vals, bio12 = bio12_vals),
    types    = c("linear", "quadratic", "hinge"),
    n_hinges = 15
)

cat(length(features), "features generated from 2 variables\n")
```

## The Optimizer

Maxent uses **sequential coordinate ascent** (also called sequential
updating) to maximize the penalized log-likelihood.
In each iteration, the algorithm:

1. Selects one feature $f_j$.
2. Finds the optimal step size $\alpha$ that maximizes the gain
   (regularized log-likelihood improvement) for that feature.
3. Updates $\lambda_j \leftarrow \lambda_j + \alpha$.
4. Recalculates the model distribution.

This process repeats, cycling through all features, until convergence
(gain per iteration falls below a threshold, typically $10^{-5}$).

### Regularization

Without regularization, Maxent can overfit to noise in the training data.
The algorithm applies **L1 regularization** (lasso penalty) controlled by
the `beta_multiplier` parameter:

- `beta_multiplier = 1.0` (default): standard regularization as in Java
  Maxent.
- `beta_multiplier > 1.0`: stronger regularization, simpler models.
- `beta_multiplier < 1.0`: weaker regularization, more complex models.

The regularization parameter for each feature depends on its type.
Hinge and threshold features receive higher penalties than linear features,
reflecting the greater risk of overfitting with more flexible feature types.

## Output Transformations

The trained model produces **raw** output values (unnormalized Gibbs
scores).
These are transformed into ecologically interpretable quantities using one
of three transformations:

### Raw Output

$$
\text{raw}(x) = \frac{\exp\left(\sum_j \lambda_j f_j(x)\right)}{Z}
$$

where $Z = \sum_x \exp\left(\sum_j \lambda_j f_j(x)\right)$ is the
density normalizer (sum over all background cells).

Raw output represents **relative occurrence rate** (ROR).
A cell with raw value 2.0 has twice the expected occurrence rate of a cell
with value 1.0.

```r
pred_raw <- maxent_project_raw(model, grids, feature_names)
```

### Cloglog Output (Recommended)

$$
\text{cloglog}(x) = 1 - \exp\left(-e^H \cdot \text{raw}(x)\right)
$$

where $H$ is the model entropy.

Cloglog output is the **recommended default** since Maxent 3.4.0.
Under certain assumptions about the sampling process, cloglog values can
be interpreted as the probability that the species is present in a grid
cell, conditional on the environment.
Values are in [0, 1].

```r
pred_cloglog <- maxent_project_cloglog(model, grids, feature_names)
```

### Logistic Output

$$
\text{logistic}(x) = \frac{e^H \cdot \text{raw}(x)}{1 + e^H \cdot \text{raw}(x)}
$$

Logistic output was the default before Maxent 3.4.0.
Like cloglog, values are in [0, 1], but the probabilistic interpretation
requires a stronger assumption (species prevalence = 0.5).
The cloglog transform is now preferred for most applications.

```r
pred_logistic <- maxent_project_logistic(model, grids, feature_names)
```

## Model Diagnostics

### AUC (Area Under the ROC Curve)

AUC measures the model's ability to distinguish presence from background
locations.
An AUC of 0.5 indicates random performance; values above 0.7 are generally
considered useful, and above 0.8 good.

### Permutation Importance

For each variable, the values at presence locations are randomly shuffled
and the AUC is recalculated.
The drop in AUC indicates how much the model depends on that variable.
Permutation importance is more reliable than percent contribution for
assessing variable effects, because it is independent of the path taken
during training.

### MESS (Multivariate Environmental Similarity Surface)

MESS maps identify areas where the model is projecting into novel
environments not represented in the training data.
Negative MESS values indicate extrapolation --- predictions in those areas
should be interpreted with caution.

### Clamping

Clamping restricts environmental variable values to their training range
before projection.
This prevents the model from extrapolating feature values beyond the
range it was trained on, producing more conservative (but safer)
predictions.

```r
clamped_grids <- maxent_clamp(grids, feature_names, occ_rows, occ_cols)
```

## Further Reading

- Phillips, S.J., Anderson, R.P. & Schapire, R.E. (2006). Maximum entropy
  modeling of species geographic distributions. *Ecological Modelling*,
  190, 231--259.
  [doi:10.1016/j.ecolmodel.2005.03.026](https://doi.org/10.1016/j.ecolmodel.2005.03.026)

- Elith, J., Phillips, S.J., Hastie, T., Dudík, M., Chee, Y.E. & Yates,
  C.J. (2011). A statistical explanation of MaxEnt for ecologists.
  *Diversity and Distributions*, 17, 43--57.
  [doi:10.1111/j.1472-4642.2010.00725.x](https://doi.org/10.1111/j.1472-4642.2010.00725.x)

- Merow, C., Smith, M.J. & Silander, J.A. (2013). A practical guide to
  MaxEnt for modeling species' distributions: what it does, and why inputs
  and settings matter. *Ecography*, 36, 1058--1069.
  [doi:10.1111/j.1600-0587.2013.07872.x](https://doi.org/10.1111/j.1600-0587.2013.07872.x)
